TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-06 src/paper.tex
update

\begin{table*}
    \caption{The performance of different systems on Libri2Mix testset. For TSELM-L-Hybrid  
    and TSELM, we compare the speaker similarity with the discretized target speech 
    (Target-discrete-WavLM) instead of the target speech due to the observation that 
    the process of discretization already losses the speaker information (0.653 for 
    Target-discrete-WavLM). We use "\_d" to denote it.  }
    \setlength{\tabcolsep}{12pt} % Adjust column spacing
    \renewcommand{\arraystretch}{1.2}
    \begin{center}
        \begin{tabular}{ccccccccc}
            \Xhline{2\arrayrulewidth} % Bold top line
            \multirow{2}{*}{System} & \multirow{2}{*}{Category} & \multirow{2}{*}{Model type} & \multirow{2}{*}{Model size} & \multicolumn{3}{c}{ DNSMOS $\uparrow$} & \multirow{2}{*}{dWER $\downarrow$} & \multirow{2}{*}{Spk Sim $\uparrow$} \\
            \cline{5-7}
              &                   &                             &                             & SIG     & BAK     & OVL    &                       &                          \\ 
            \hline
            Mixture                 & - & -                           & -                           & 3.39    & 3.14    & 2.68   & 79.3\%                & -                        \\
            Target-Discrete   & G & -                          & 330M                           & 3.57    & 4.10    & 3.32   & 11.3\%                & 0.653                    \\
            \hline
            Spex+                   &  D & Continuous                  & 11M                         & 3.36    & 3.76    & 2.98   & 19.3\%                & 0.923                    \\
            \hline
            \hline
            Continuous-WavLM-L6   & G    & Continuous                  & 220M                        & 3.56    & 4.06    & 3.27   & 14.7\%                & 0.877                    \\
            TSELM-L-Hybrid    & G       & Hybrid                      & 468M                        & 3.50    & 4.06    & 3.22   & 19.8\%                & 0.924\_d                 \\
            TSELM-S-NoConcat           & G       & Discrete                    & 354M                        & 3.47    & 4.03    & 3.19   & 69.6\%                & 0.868\_d                 \\
            \hline
            TSELM-S           & G       & Discrete                    & 354M                        & 3.50    & 4.07    & 3.24   & 28.1\%                & 0.892\_d                 \\
            TSELM-M          & G        & Discrete                    & 389M                        & 3.49    & 4.05    & 3.22   & 28.4\%                & 0.901\_d                 \\
            TSELM-L         & G         & Discrete                    & 468M                        & 3.49    & 4.05    & 3.22   & 27.1\%                & 0.905\_d \\
            \Xhline{2\arrayrulewidth} % Bold top line              
            \end{tabular}
            \linebreak
            \label{main_exp}
      \end{center}
    %   \vspace{-10pt}
  \end{table*}

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. Since discretization can potentially lose speaker information compared to continuous methods. We compute the speaker 
similarity of output from TSELM with the target audio produced by discretizing the clean audio, denoted by Target-Discrete, representing the upper bound of our model  
performance. Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoConcat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross attention  
and LM without discretization. 
The concatenation strategy is stilled applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoConcat, we utilize the TSELM-S model architecture but without concatenation strategy.

\section{Results and Discussions}
Table \ref{main_exp} shows the performance of different systems on the Libri2Mix test.
The model size for Target-discrete-WavLM denotes the model size of the encoder and 
decoder, which is WavLM Large and HiFi-GAN respectively. DNSMOS is calculated over the 
test sets since it is reference-free. dWER is calculated with the clean speech instead 
of the discretized speech due to our observations that this does not make much difference. 
For TSELM-L-Hybrid and TSELM, we compute the speaker similarity with the discretized
target speech (Target-discrete-WavLM) instead of the target speech.
% due to the observation that the process of discretization already losses the speaker
% information (0.653 for Target-discrete-WavLM). 
We observe that the process of discretization already losses the speaker
information (0.653 for Target-discrete-WavLM), and it might be related to the inevitable speaker 
information loss by the process of discretization. Future work should focus on developing better 
tokenization methods for SSL models. 
Since work focuses on using the discretized information 
to do target speaker extraction instead of developing better tokenization methods, we think it is 
reasonable to compare our outputs with the discretized speech since it serves as an upper bond.

We observe that our model performs better than Spex+ in terms of the DNSMOS scores but worse in dWER. 
This suggests that our model has better speech quality compared with discriminative models but 
worse in speech intelligibility. One possible reason might be related 
to the discretization process on the mixture 
speech. Our Kmeans algorithm is trained on clean speech instead of mixed speech. 
For speech enhancement, this might be 
good because the level of noise will usually be smaller than the speech. Kmeans might achieve the 
effect of denoising. Discretization on the mixed
speech, however, might cause the output to focus on wrong speakers. 
Our results from TSELM-L-Hybrid 
support 
this. For TSELM-L-Hybrid, we use the continuous embeddings from the mixture speech without discretizing, and it has 
achieved similar dWER with Spex+. Another reason might be related to LM. Right now our encoder-only 
LM satisfies roughly 50\% accuracy. We think using auto-regressive or masking models will achieve 
better results.

We observe significant dWER increasing without concatenating the reference audio to the mixture, 
as shown 
the TSELM-S-NoConcat in Table \ref{main_exp}. WavLM \cite{wavlm} is demonstrated to have target speaker 
separation capability. The input audio should have the mixture less than 50\% and have the first utterance as target speaker. The output will be a coarse denoised embedding emphasizing the target speaker. 
If the whole input is mixture, we have found that WavLM sometimes will 
extract the wrong speaker. Inspired by SELM's \cite{selm} success in speech denoising. Our pipeline is trying to transform
the target speaker separation problem to speech enhancement problem by using WavLM's denoising capability.

In Table \ref{hubert_wavlm_6}, we have compared the performance of different SSL models and the selection between one layer or multiple layers to discretize. 
Our results have shown that the using HuBERT as SSL model has achieved better DNSMOS scores but with worse dWER compared to our WavLM baseline.
The better DNSMOS scores might be related to the vocoders but it does not reflect the speech intelligibility which is important in speech separation. 
For dWER, this might be because that HuBERT is trained on clean speech and that it might not capture the rich speech information of mixed speeches. 
Our results from WavLM-L6 have shown that for speech separation, using multiple layers to discretize is better than using only one layer.

Finally, we have observed a performance gap between discrete methods and continuous methods (Continuous-WavLM-L6) on target 
speaker extraction as shown in Table \ref{main_exp}. For Continuous-WavLM-L6,
we use the 6th layer embedding output from WavLM from both the mixed speech and reference speech, and use MSE loss between the output from LM and the clean embeddings. This method 
has the best performance in terms of DNSMOS and dWER. It has a better dWER of 13.6\% compared with TSELM-L. This might be related to the 
information loss of discretization process. We hope future research will work to bridge this gap.

\section{Conclusion}
In this work, we introduced a novel way using discrete tokens and language models for target speaker extraction.
We use multiple hidden layers of WavLM and Kmeans tokenizers as encoder, cross attention and language model as our separation model, and a scalable HiFi-GAN to reconstruct the 
audio. 
Experiments have 
shown that our model can achieve excellent performance in terms of speech quality, and comparable performance in terms of speech intelligibility and 
speaker similarity. However, we still have a gap between continuous methods especially in speech intelligibility and speaker similarity. Future research should focus on shrinking 
this gap.

>>>>

\begin{table*}
    \caption{The performance of different systems on Libri2Mix testset. "G" means generative while "D" means discriminative. For discrete models,
    we compute the speaker similarity with the discretized target speech 
    (Target-discrete-WavLM). We use "\_d" to denote it.  }
    \setlength{\tabcolsep}{12pt} % Adjust column spacing
    \renewcommand{\arraystretch}{1.2}
    \begin{center}
        \begin{tabular}{ccccccccc}
            \Xhline{2\arrayrulewidth} % Bold top line
            \multirow{2}{*}{System} & \multirow{2}{*}{Category} & \multirow{2}{*}{Model type} & \multirow{2}{*}{Model size} & \multicolumn{3}{c}{ DNSMOS $\uparrow$} & \multirow{2}{*}{dWER $\downarrow$} & \multirow{2}{*}{Spk Sim $\uparrow$} \\
            \cline{5-7}
              &                   &                             &                             & SIG     & BAK     & OVL    &                       &                          \\ 
            \hline
            Mixture                 & - & -                           & -                           & 3.39    & 3.14    & 2.68   & 79.3\%                & -                        \\
            Target-Discrete   & G & -                          & 330M                           & 3.57    & 4.10    & 3.32   & 11.3\%                & 0.653                    \\
            \hline
            Spex+                   &  D & Continuous                  & 11M                         & 3.36    & 3.76    & 2.98   & 19.3\%                & 0.923                    \\
            \hline
            \hline
            Continuous-WavLM-L6   & G    & Continuous                  & 220M                        & 3.56    & 4.06    & 3.27   & 14.7\%                & 0.877                    \\
            TSELM-L-Hybrid    & G       & Hybrid                      & 468M                        & 3.50    & 4.06    & 3.22   & 19.8\%                & 0.924\_d                 \\
            TSELM-S-NoConcat           & G       & Discrete                    & 354M                        & 3.47    & 4.03    & 3.19   & 69.6\%                & 0.868\_d                 \\
            \hline
            TSELM-S           & G       & Discrete                    & 354M                        & 3.50    & 4.07    & 3.24   & 28.1\%                & 0.892\_d                 \\
            TSELM-M          & G        & Discrete                    & 389M                        & 3.49    & 4.05    & 3.22   & 28.4\%                & 0.901\_d                 \\
            TSELM-L         & G         & Discrete                    & 468M                        & 3.49    & 4.05    & 3.22   & 27.1\%                & 0.905\_d \\
            \Xhline{2\arrayrulewidth} % Bold top line              
            \end{tabular}
            \linebreak
            \label{main_exp}
      \end{center}
    %   \vspace{-10pt}
  \end{table*}

@@

\subsection{Baseline models}
Baseline models are presented in Table \ref{main_exp}. 
TSELM is compared with Spex+ \cite{spex_plus}, a discriminative separation model trained on Libri2Mix \cite{librimix}. 
Mixture refers to the unprocessed mixed speech. 
Target-Discrete refers to the discretized target speech. It serves as the upper bond for our model.  Besides TSELM, we conduct three main experiments, named
 Continuous-WavLM-L6, TSELM-L-Hybrid and TSELM-S-NoConcat using the same training data. 
For Continuous-WavLM-6, we directly
pass the embeddings from the 6th hidden layer output of WavLM Large to the cross attention  
and LM without discretization. 
The concatenation strategy is stilled applied to the mixed speech. Mean Square Error (MSE) loss is applied between the output embeddings and the clean 
embeddings. HiFi-GAN in \cite{knn_vc} is used for audio reconstruction. For 
TSELM-L-Hybrid, inspired by MaskSR \cite{mask_sr}, we discretize the reference 
speech while utilizing the continuous embeddings from the mixed speech. We call 
it hybrid because the mixture speech retains continuous features. In TSELM-S-NoConcat, we utilize the TSELM-S model architecture but without concatenation strategy.

\section{Results and Discussions}
Table \ref{main_exp} presents the performance of different systems evaluated on the Libri2Mix test.
The model size for Target-Discrete refers to the total size of the encoder and 
decoder, which are WavLM Large and scalable HiFi-GAN, respectively. DNSMOS is computed over the 
testset since this metric is reference-free. dWER is calculated with respect to the clean speech. 
For continuous methods, speaker similarity is directly computed against the clean speech. 
However, since discretization inherently results in some loss of speaker information—Target-Discrete shows a speaker similarity score of 0.653 compared to continuous methods—we assess the speaker similarity of the output from the discrete methods against the target audio produced by discretizing the clean speech. 
This target audio output from Target-Discrete serves as the upper bond for our model performance.
The observed speaker information loss is likely due to the tokenization process, which inherently reduces speaker fidelity. Future work should aim to enhance tokenization methods for SSL models to mitigate this loss. Since the primary goal of this research is to explore target speaker extraction using discretized information rather than to develop improved tokenization methods, we argue that comparing the outputs of our model with discretized speech is a reasonable approach, as it represents an upper bound on performance.

We observe that our model outperforms Spex+ in terms of DNSMOS scores, indicating better speech quality, but performs slightly worse in dWER, suggesting lower speech intelligibility. One potential reason for this could be the discretization process applied to the mixed speech. Our Kmeans algorithm is trained on clean speech rather than mixed speech, which is advantageous for speech enhancement as it likely aids in denoising. However, when applied to mixed speech, this discretization might lead to the model focusing on the wrong speaker, as it may only retain dominant speaker information, causing a reduction in intelligibility.
This hypothesis is supported by the results of TSELM-L-Hybrid, where continuous embeddings from the mixture speech were used without discretization, achieving dWER scores similar to Spex+. Another contributing factor could be the limitations of our current language model (LM). The encoder-only LM achieves around 50\% accuracy, and we believe using more advanced models, such as auto-regressive or masking-based LMs, could lead to improved performance in future iterations.

We observe a significant increase in dWER when the mixture audio is not concatenated with the reference, as seen in the TSELM-S-NoConcat results in Table \ref{main_exp}. For WavLM to effectively perform target speaker separation, the input audio must follow specific conditions: the mixture should be less than 50\% of the total length, and the first utterance should be the target speaker. Under these conditions, WavLM outputs a slightly denoised embedding that emphasizes the target speaker. When the entire input is a mixture, however, we found that WavLM sometimes extracts the wrong speaker.
Our current concatenation strategy is inspired by SELM's \cite{selm} success in speech denoising and aims to transform the target speaker extraction task into a more difficult speech enhancement problem leveraging WavLM's denoising capability.

In Table \ref{hubert_wavlm_6}, we compare the performance of HuBERT and WavLM as SSL model and examine the effects of using either one or multiple layers for discretization. Our findings indicate that using HuBERT as the SSL model results in better DNSMOS scores but much worse dWER compared to our WavLM baseline. The improved DNSMOS scores likely stem from the vocoder performance, yet they do not adequately reflect speech intelligibility, which is crucial for speech separation tasks. 
The poorer dWER scores observed with HuBERT may be attributed to its training on clean speech, which might not equip it to capture the complexity and richness of mixed speech. Moreover, our results from WavLM-L6 suggest that when performing speech separation, discretizing across multiple layers provides better results than relying on a single layer.

Finally, we observed a performance gap between discrete methods and continuous methods, as demonstrated by Continuous-WavLM-L6.
Continuous-WavLM-L6 has best performance in terms of DNSMOS and dWER among all the experiments.
It has a better dWER of 13.6\% compared with TSELM-L. The gap in performance may be attributed to the information loss inherent in the discretization process. 
We hope future research will work to bridge this gap.

\section{Conclusion}
In this work, we introduced a novel way using discrete tokens and language models for target speaker extraction.
Our method leverages multiple hidden layers of WavLM and Kmeans tokenizers for encoding, employs cross-attention and a language model for separation, and utilizes a scalable HiFi-GAN for audio reconstruction.
Experiments have 
shown that our model can achieve excellent performance in terms of speech quality, and comparable performance in terms of speech intelligibility and 
speaker similarity. However, we still have a gap between continuous methods especially in speech intelligibility and speaker similarity. Future research should focus on shrinking 
this gap.
